---
title: "Chapter 9: Git and GitHub"
author: "Lorraine Gaudio"
date:   "`r paste('Version', format(Sys.Date(), '%B %d, %Y'))`"
team: "Spring 2026"
output: 
  html_document: # To create an HTML document from R Markdown
    toc: false # Table of contents (TOC)
    toc_depth: 1 #(meaning that level 1, 2, and 3 headers will be included in the table of contents
    toc_float: # Float the table of contents to the left of the main document
      collapsed: false # Collapsed (defaults to TRUE) controls whether the TOC appears with only the top-level
      smooth_scroll: true # controls whether page scrolls are animated when TOC items are navigated to via mouse clicks.
    number_sections: true # Numbering starts with "#" (H1). Without H1 headers, the H2 headers ("##") will be numbered with 0.1, 0.2, and so on.
    css: ../assets/styles.css # This is the name of the CSS file to style the HTML document with Boise State Brand. The CSS file must be in the same directory as the R Markdown file.
    fig_caption: true #Whether figures are rendered with captions.
    df_print: paged # Printing data frames with interactivne scrolling
    includes:
      in_header: ../assets/header.html
      after_body: ../assets/footer.html
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# Overview

*This is both Module 4 and 5*

TODO: WRITE AN OVERVIEW AND WHY LEARNING THE CHAPTER CONTENT IS IMPORTANT.

![](https://intro2r.com/github_r.html)

In Chapter 9 you will learn how to:

- Courage Skill Highlight

Students can attempt a task first (even imperfectly), then seek help with a clear question and evidence (code + error + what they think is happening).

Students can share imperfect work in a discussion/lab setting and revise based on feedback.

- Git

- Github

- RStudio Projects

- Workflow

![Happy git with R](https://happygitwithr.com/)

## What is Git?

**Git**: a collection of individual commands you execute in the shell

- Version control system

- Help collaborate in a group

- Tracks changes

Repository: a set of files stored by a provider

- GitHub is a repository hosting service for your Git-based projects online.

- Allows others access to your project in nearly real time

- Remote storage

- Access and edit locally

### Privacy and Github

Covers (short, concrete):

- Why `data/` exists: so later you can ignore it when you start using GitHub.

- The mechanism they’ll use later: a .gitignore can exclude folders like `data/` from commits.

- A simple rule of thumb:

1. OK to upload: public/open data, de-identified data you’re allowed to share

2. Do not upload: human-subjects data, protected/controlled data, anything you promised to keep private



## Plan

Part 1: Initial Set Up

1.  Install Git

2. Get Git talking to GitHub

3. Make RStudio talk to your local Git


💡 **Tip:** Merge conflicts issues happen if you don't *always* "Pull" first.

Part 2: New Projects Workflow

1. Create a repo folder in GitHub

2. Make RStudio Project

3. Make is a Git repository

4. Pull to access latest version

5. Periodically make a commit

6. Push To GitHub

# Part 1

Account set up and learning Git

1. Register for free GitHub Account

2. Install or update R and Studio

3. Install Git

4. Introduce yourself to Git

5. Configure a SSH keys

## Register for free GitHub Account

The first step is to Register a GitHub Account.

Your username tips:

- short, 

- professional, 

- memorable, 

- timeless, 

- all lower case, 

- separate words with a hyphen “-”

## Install or update R and Studio

Make sure you have the latest versions of R and RStudio installed. This will ensure compatibility with Git and GitHub.

![Download R and RStudio](https://posit.co/download/rstudio-desktop/)

## Install Git

### Windows: Git Bash

![gitforwindows](https://gitforwindows.org/)

1. Accept the defaults

2. RStudio for Windows prefers for Git to be installed below C:/Program Files

    C:/Program Files/Git/bin/git.exe

### macOS: Xcode command line tools

1. Go to the shell and enter one of these commands to elicit an offer to install developer command line tools:

    git --version

    git config

2. Accept the offer, click install

OR

![git scm mac](http://git-scm.com/downloads)

*Dr. John Ziker revised instructions*

1. Download Xcode from the App Store

2. Use “Terminal” app on Mac for next step

### Linux: distro’s package manager

1. Ubuntu or Debian Linux: sudo apt-get install git

2. Fedora or RedHat Linux: sudo yum install git

OR

![git scm (linux)](https://git-scm.com/download/linux)

## Introduce yourself to Git

Introduce yourself to Git in shell

    git config --global user.name "Your Name"
    git config --global user.email "example@u.boisestate.edu"

*Use the email associated with your GitHub account*

Check Git understood

    git config --global --list

Initial default branch

  git config --global init.defaultBranch main

GitHub has “main” as the default initial branch 

## Configure SSH keys

Connecting Tools

Personal Access Token (PAT)

OR

SSH Keys

PATs are convenient but easily compromised.

SSH Keys never leave your machine, allow for more control and provides better security, but are more challenging to set up. 

![Mac users instructions](https://gist.github.com/brennanMKE/8e09593ca4064deab59da807077d8f53) for creating an SSH key.

### Setup SSH 

**SSH keys**: two long strings of characters used to authenticate a user’s identity when connecting to a remote server. 
Generate these keys on your local computer using an SSH utility.

1. From the shell: `ssh-keygen -t rsa -b 4096`

2. Follow prompts to create your keys. By default, they’ll be saved in `~/.ssh/id_rsa` (private key) and `~/.ssh/id_rsa.pub` (public key).

3. Add your public key to GitHub

a. Go to settings > SSH and GPG keys

b. Click “New SSH key”

c. Copy public key into key field

d. Give device a title

4. Test Connection

a. From the shell: `ssh -T git@github.com`

1. Randomness and Key Generation:

- The ssh-keygen command generates an RSA key pair (consisting of a private key and a public key).

- The process starts by creating a random seed (using entropy sources like keyboard timings and mouse movements).

- This randomness is essential for security.

2. Mathematical Algorithms:

- The RSA algorithm is used to create the key pair.

3. The **private key** must remain confidential and secure. It is used for authentication.

4. You Never Share Your Private Key

# Part 2

Prove local Git can talk to GitHub

1. Make a Repo on GitHub

2. Clone the Repo in Shell

3. Make a local change

## Make a Repo on GitHub

1. Go to ![GitHub.com](https://github.com) and make sure you are logged in.

2. Near “Repositories”, click the big green “New” button. 

3. Repository template: No template

4. Repository name: myrepo

5. Description: “My first repo”

6. Public

7. Initialize: Add a README file

8. Click green “Create repository”

9. In Local tab, select HTTPS SSH

10. Copy

## Clone Repo in Shell

1. Paste Copied: 

    git clone https://github.com/YOUR-USERNAME/YOUR-REPOSITORY.git

2. Make working directory

    cd myrepo
    ls
    head README.md
    git remote show origin

## Make a local change

1. Add line to README


    echo "A line I wrote on my local computer  " >> README.md
    git status


2. Send to GitHub

    git add README.md
    git commit -m "A commit from my local computer"
    git push
    
💡 **Tip:**When finished, delete folder corresponding to the local repo

    cd ..
    rm -rf myrepo/


# Part 3

Connecting Tools: Connect RStudio to GitHub

Prove RStudio can talk to GitHub

1. Make a repo on GitHub

2. Clone GitHub via RStudio

3. Make Local changes, save and commit

4. Confirm the local change in GitHub

## Make a repo on GitHub

1. Go to ![GitHub](https://github.com) and make sure you are logged in.

2. Near “Repositories”, click the big green “New” button. 

3. Repository template: No template

4. Repository name: `myrepo`

5. Description: “My second repo”

6. Public

7. Initialize: Add a README file

8. Click green “Create repository”

9. In Local tab, select HTTPS SSH

10. Copy repository URL

## Clone GitHub via RStudio

Start a new project

1. File > New Project > Version Control > Git

2. Paste repository URL

3. Accept the default (project name and GitHub repo name must match)

4. PAY ATTENTION to where you are saving your project locally

5. Open in a new session

6. Create Project

7. Notice the README.md file on your computer

## Local changes

1. Open README.md in RStudio

2. Add line to README

    “A line I wrote in RStudio"

3. Commit

a. Click Git tab

b. Check “Staged” box for README.md

c. Click “Commit”

d. Type message “Commit from R”

e. Click Commit

4. Push

5. Confirm in GitHub

# Part 4

Workflow

## Create a R Markdown Document

File > New File > R Markdown

1. Make sure your project is open

2. Give it an informative title (Spaces and Punctuation)

3. Output format of HTML (Can change later)

4. Save “first.rmd” in RStudio project folder (Do not nest in folder)

No need to set a working directory again!

## Edit the R Markdown Document

1. Add some text and code

2. Save

3. Commit 

4. Push to GitHub

## Data Folder Structure

Covers:

1.. Minimum viable structure (what you described): keep the data in the same folder as the R Notebook so relative paths are simple.

2. Best practice structure: create a `data/` folder and move raw files into it:

-`Intro_to_R/`
-- `chapter9_notes.Rmd`
-- `data/`
--- `myfile.csv`
--- `myfile.xlsx`

Introduce RStudio Projects: a way to keep everything together and keep a stable working directory.

### Privacy and Github

Covers (short, concrete):

- Why `data/` exists: so later you can ignore it when you start using GitHub.

- The mechanism they’ll use later: a .gitignore can exclude folders like `data/` from commits.

- A simple rule of thumb:

1. OK to upload: public/open data, de-identified data you’re allowed to share

2. Do not upload: human-subjects data, protected/controlled data, anything you promised to keep private


gitignore: a file that tells Git which files or folders to ignore when committing changes. This is useful for keeping sensitive data or large files out of your GitHub repository.

In Chapter 7, you learned to store data in the same folder as your R Notebook. This is the minimum viable structure for a project, but it can get messy as you add more files. A better practice is to create a `data/` folder and move your raw data files into it. This way, your project folder stays organized, and you can easily manage your data files.

Your structure becomes:

-`Intro_to_R/`
-- `chapter9_notes.Rmd`
-- `data/`
--- `myfile.csv`
--- `myfile.xlsx`

**🎯 Goal** Store your data in a `data/` folder then check 

```{r}
# Check your R notebook's working directory files
list.files()
list.files("data")
```

🗣 After moving the files into data/, you must include the `data/` in the file path.

# Summary


# Chapter Terms


# 📝 Practice Space

## Task 1

## Task 2

# References


