---
title: "Chapter 5 Practice"
output: html_document
---

## Task 1

🎯: Practice creating vectors and doing arithmetic. Understand R syntax.

```{r}
#  ⚡ 
result <- 3+4 
# number <,-42
# ratio <- number / result
```

Error: unexpected ',' in "number <,"
Error: object 'number' not found

✍ Memo:  I accidentally had "," in there which it didn't recognize. 
Then it couldn't run this because it didn't have a value for "number" yet. 

```{r}
#  ⚡  
number <- 42

ratio <- number / result

# print(Ratio)
```

Error: object 'Ratio' not found

✍ Memo: cases matter

```{r}
# ✅ Check the ratio value
print(ratio)
```

🗣 R successfully calculated the ratio of `number` to `result`, which is 42 divided by 7, resulting in 6.

## Task 2

🎯: Create two vectors of different lengths and subtract them to see how recycling works.

```{r eval=FALSE}
# ⚡ 
long <- c(6, 2, 8, 10, 3, 77)
short <- c(88, 4, 20)
difference <- long - short
# ✅ 
print(difference)
```

✍ Memo: I struggled to understand how R recycles vectors of different lengths during arithmetic operations. I asked ChatGPT what the pattern was, and it said the results are [1] -82  -2 -12 -78  -1  57 which is essentially 6-88, 2-4, 8-20, 88-10, 3-4, 77-20

🗣 When subtracting a vector of unequal lengths, R recycles the shorter vector to match the length of the longer one.

## Task 3


🎯 How does the `paste()` function work?

```{r eval=FALSE}
# ⚡ 
first_word<-"yellow"

second_word<-"banana"
sentence<-paste(first_word,second_word)
sentence
```

✍ Memo: I wasn't expecting it to turn yellow!!! The sentence came out as "yellow banana"...neat!

```{r eval=FALSE}
class(sentence)
```

✍ Memo: Character...? Uh, what's that referring to??


## Task 4

🎯 Make a vector of class standings and then convert it to factor class.

```{r}
# ⚡
standing<-c("Freshman","Sophomore","Junior","Senior","Freshman","Freshman")
# ✅ 
standing
class(standing)
```

🗣 okay so it prints the words. It says class is character.

```{r}
# ⚡
standing_f<-factor(standing)
# ✅ 
standing_f
```
```{r}
# ✅
mode(standing_f)
typeof(standing_f)
levels(standing_f)
```

✍ Memo:  The `mode()` says numeric, `typeof()` says integer?? but it PRINTS WORDS.
I get that levels shows the categories. How is it storing numbers and then showing character labels?? This is confusing!

👀 Memo:  I asked ChatGPT to explain  why a character vector changed to a factor class is stored as integer type, numeric mode and codes with text labels called levels. It clarified that factors are stored as integer codes with a separate set of text labels called levels, which explains why `typeof()` is "integer" even though printing shows words. It also confirmed that `levels()` displays the labels.


