---
title: "Chapter 12 misc"
author: "Lorraine Gaudio"
date: "2026-03-17"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Overwrite with Factors

🎯 Goal: Take a 100-row sample from `movielens`, then overwrite the numeric `rating` column with an ordered factor: "Bad" if `< 4`, "Good" if `>= 4`, using `case_when()` inside `mutate()`.

💬 Prediction: Before: `rating` is numeric. After: `rating` is an ordered factor with levels `Bad < Good`. Row count stays 100; no `NA`s created.

📜 SYNTAX reminder:
`mutate(data, new_col = case_when(cond1 ~ value1, cond2 ~ value2, TRUE ~ default))`

`name_data <- data %>% mutate(new_col = case_when(cond1 ~ value1, cond2 ~ value2, TRUE ~ default))`

```{r , eval=FALSE}
data("movielens")
?movielens
```

```{r , eval=FALSE}
set.seed(99)

# Sample 100 movies from the movielens dataset
smaller_movies <- movielens %>% 
  sample_n(100) %>% 
  select(title, rating)

head(smaller_movies)
```

```{r , eval=FALSE}
# ✅ Checks (before)
names(smaller_movies)                # expect "title", "rating"
class(smaller_movies$rating)         # expect "numeric"
summary(smaller_movies$rating)       # quick distribution check
```

🎯 Goal: We'll apply `mutate()` to `smaller_movies` to recode the rating column to "Good" or "Bad".

```{r , eval=FALSE}
# ⚡ Overwrite numeric rating with an ordered factor via case_when()
movies_overwrite <- smaller_movies %>%
  mutate(
    rating = factor(
      case_when(
        rating >= 4 ~ "Good",
        rating < 2 ~ "Bad",
        TRUE        ~ "Ok"
      ),
      levels = c("Bad", "Ok", "Good"),
      ordered = TRUE
    )
  )
```


```{r , eval=FALSE}
# ✅ Checks (after)
names(movies_overwrite)              # still "title", "rating"
```


```{r , eval=FALSE}
# ✅ Checks (after)
is.ordered(movies_overwrite$rating)  # expect TRUE
```


```{r , eval=FALSE}
# ✅ Checks (after)
levels(movies_overwrite$rating)      # expect "Bad", "Ok", "Good"
```


```{r , eval=FALSE}
# ✅ Checks (after)
table(movies_overwrite$rating)       # counts in each level
```

🗣 Comment: Overwriting keeps the column name but replaces its contents and type (numeric → ordered factor). 

🪞 Metacognition: `case_when()` reads cleaner than nested `ifelse()` for multi-branch recodes and makes me specify a default. In the future, I could keep the original numeric ratings (e.g., `rating_num`) and add a new labeled column instead of overwriting, to preserve provenance.


```{r , eval=FALSE}
# ⚡ Overwrite numeric rating with an ordered factor via case_when()
movies_rename <- smaller_movies %>%
  mutate(
    rating_words = factor(
      case_when(
        rating >= 4 ~ "Good",
        rating < 2 ~ "Bad",
        TRUE        ~ "Ok"
      ),
      levels = c("Bad", "Ok", "Good"),
      ordered = TRUE
    )
  )
```

### Bonus Example

🎯 Goal: Create a new column rating_label in smaller_movies using `case_when()` that labels ratings `>= 4` as "Good" and the rest as "Bad". Keep the original numeric rating column.

💬 Prediction: After running, the data will still have 100 rows, keep both `title` and `rating`, and add `rating_label` (character or factor). No `NA`s should appear.

```{r , eval=FALSE}
# ✅ Checks (before)
names(smaller_movies)                # expect "title", "rating"
class(smaller_movies$rating)         # expect "numeric"
```

```{r , eval=FALSE}
# ⚡ Add a NEW column using case_when()
smaller_movies %>%
  mutate(
    rating_label = case_when(
      rating >= 4 ~ "Good",
      TRUE        ~ "Bad"
    )
  ) -> movies_newcol
```

🧐 NOTICE: What is happening here `-> movies_newcol`?

🛠Break Things! How can you reorganize this script? 

```{r , eval=FALSE}
# ✅ Checks (after)
names(movies_newcol)                 # expect "title", "rating", "rating_label"
class(movies_newcol$rating)          # expect "numeric"
class(movies_newcol$rating_label)    # expect "character" or "factor"
```
🗣 The code used -> movies_newcol, which assigns right-to-left.

🛠 Break Things! What happens with `head(movies_newcol) -> movies_newcol`?

🪞 Metacognition: I’ll stick with the usual `<-` to make my code easier to read

## Calculations

🚗 Using `mtcars`, let's create a column `l_per_100km` that converts from US units (miles per gallon) to metric units (liters per 100km). 235.215 is the conversion factor between these units

🎯 Goal: Identify the most fuel-efficient car in L/100 km terms.

💬 Prediction: The new column `l_per_100km` will be added to `mtcars_units`, and the most fuel-efficient car will have the lowest value in this column.


```{r , eval=FALSE}
# ⚡ Code
mtcars_units <- mtcars %>%
  mutate(l_per_100km = round(235.215 / mpg, 1))
```

```{r , eval=FALSE}
# ✅ Check
head(mtcars_units)
```

👀 Check‑in: Which car is most fuel‑efficient in l/100 km terms?

```{r , eval=FALSE}
# Find the most fuel-efficient (smallest L/100km)
best_idx <- which.min(mtcars_units$l_per_100km)
best_car <- rownames(mtcars_units)[best_idx]
best_value <- mtcars_units$l_per_100km[best_idx]
```

```{r , eval=FALSE}
"best car"
best_car
```

```{r , eval=FALSE}
"best value"
best_value
```

🪞 Metacognition: Unit conversions are easy to keep alongside original units with `mutate()` (don’t overwrite).
